---
title: "Gov 51 - Pset 4"
author: "Tyler Dang"
format: pdf
editor: visual
---

## Setup

```{r}
library(tidyverse)
library(estimatr)
library(modelsummary)
card <- read_csv("/Users/tyler/Downloads/card1995.csv")
```

1.2

```{r}
nrow(card)
  
```

There are 3010 observations. Each unit refers to a respondent individual male.

1.3

```{r}
library(ggplot2)

ggplot(card, aes(x = educ,)) +
  geom_histogram(binwidth = 1, boundary = 0, fill = "salmon") +
  labs(title = "Distribution of Years of Education", 
       x = "Years of Education",
       y = "Count") +
  theme_minimal()

mean(card$educ, na.rm = TRUE)
```

The average level of education is \~13 years. The graph shows distinct spikes at 12 years mark and 16 years mark which are around the ages when people graduate from high school and college. Since these spikes occur because education are represented as discrete numbers, this causes respondents to cluster at graduation endpoints rather than being uniformly distributed.

1.4

```{r}
ggplot(card, aes(x = educ, y = lwage)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Log Wages vs. Years of Education",
       x = "Years of Education",
       y = "Log Wage (lwage)") +
  theme_minimal()
```

The scatter plot shows a clear positive relationship between education and log wages. The regression line slopes upward which shows that those with more years of education tend to earn higher wages. However, there is variation in wages at each level of education which suggests that there may be other omitted factors other than education that also play an important role in predicting earnings.

1.5

Individuals with higher innate ability may both attain more education and earn higher wages which creates a positive correlation, regardless if education has any causal effect. Secondly, family background and/or socioeconomic status may influence both educational attainment and labor market outcomes which may lead to omitted variable bias in our OLS estimate.

2.1

```{r}
mean(card$nearc4, na.rm = TRUE)
```

Based on our data result, around 68.2% of respondents in the sample grew up near a 4 year college.

2.2

```{r}
first_stage <- lm_robust(educ ~ nearc4 + black + smsa + south +
                           exper + expersq + married,
                         data = card, se_type = "HC2")
first_stage
```

The coefficient on nearc4 is 0.327 with a standard error of 0.081 and a t-statistic of 4.06. This means that individuals who grew up near a 4-year college obtain 0.327 more years of education on average, holding other variables in the regression constant. The positive sign is expected as being closer to a college reduces the cost of attending through access to local scholarships and increases access to higher education.

2.3

```{r}
t_stat <- summary(first_stage)$coefficients["nearc4", "t value"]
F_stat <- t_stat^2

t_stat
F_stat
```

The t-statistic on nearc4 is 4.06. Squaring this gives a first-stage F-statistic of 16.49. Since this exceeds the threshold of 10, the instrument appears to be strong. It is important to keep in mind the strength of our instrument because weak instruments lead to biased and imprecise 2SLS estimates which makes our inference of the result to be unreliable.

2.4

**(a)** The exclusion restriction requires that, conditional on education and other controls within the regression, proximity to a college (nearc4) has no direct effect on wages. More specifically, once education is controlled for, nearc4 must be uncorrelated with the error term in the wage equation.

**(b)** One possible violation is that growing up near a college may be associated with living in more urban or economically developed areas. In which, those areas could offer better or more job opportunities and higher wages, independent of education attainment. In this case, nearc4 would affect wages through channels other than education which would violate the exclusion restriction.

2.5

```{r}
library(dplyr)

card %>%
  group_by(nearc4) %>%
  summarise(
    mean_black = mean(black, na.rm = TRUE),
    mean_smsa  = mean(smsa, na.rm = TRUE),
    mean_south = mean(south, na.rm = TRUE),
    n = n()
  )
```

Based on the output, the groups are not balanced. Individuals who grew up near a college are much more likely to live in urban areas (higher smsa) and less likely to live in the South. There are also differences in racial composition. This imbalance suggests that nearc4 is not randomly assigned and is correlated with geographic and demographic factors that may independently affect wages. As a result, we should question the validity of the instrument which relies on controlling for these variables, and the exclusion restriction.

3.1

```{r}
reduced_form <- lm_robust(
  lwage ~ nearc4 + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2")

summary(reduced_form)
```

The coefficient on nearc4 is 0.0409 with a SE = 0.0161 which indicates that individuals who grew up near a 4-year college earn around  4.1% higher wages on average, holding other factors in this regression constant. This suggests that proximity to a college is positively associated with wages even though this effect may occur both through increased education and other channels.

3.2

```{r}
rf_coef <- coef(reduced_form)["nearc4"]
fs_coef <- coef(first_stage)["nearc4"]

rf_coef / fs_coef
```

Wald Estimator: 0.1249

3.3

The Wald estimate of 0.1249 implies that one additional year of education increases log wages by about 0.125, which corresponds to about a 12.5% increase in wages.

4.1

```{r}
ols <- lm_robust(
  lwage ~ educ + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2")

summary(ols)
```

The coefficient on education is 0.0715 which implies that one additional year of education is associated with about a 7.15% increase in wages, holding other factors in the regression constant.

4.2

```{r}
iv <- iv_robust(
  lwage ~ educ + black + smsa + south + exper + expersq + married |
  nearc4 + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2")

summary(iv)

coef(iv)["educ"]
```

The 2SLS coefficient on educ is 0.1249 with SE = 0.0491. This matches the Wald estimate computed earlier (0.1249). The equality is expected because the model has a single instrument so the number of known parameter is the same as the number of unknown parameter which means it is the special case where the 2SLS and Wald estimators coincide.

4.3

```{r}
modelsummary(
  list("OLS" = ols, "2SLS" = iv),
  coef_map = c("educ" = "Years of Education"),
  gof_map = data.frame(
    raw = c("nobs", "r.squared"),
    clean = c("N", "R²"),
    fmt = c(0, 3)
  )
)
```

The OLS estimate of the return to education is 0.071, while the 2SLS estimate is 0.125. The IV estimate is significantly larger than the OLS estimate, suggesting that OLS may be biased downward, perhaps because of potential measurement error or heterogeneous returns to education. The 2SLS estimate is less precise since its SE range is larger.

4.4

```{r}
ols_iq <- lm_robust(
  lwage ~ educ + IQ + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2")

summary(ols_iq)
coef(ols_iq)["educ"]

iv_iq <- iv_robust(
  lwage ~ educ + IQ + black + smsa + south + exper + expersq + married |
  nearc4 + IQ + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2"
)

summary(iv_iq)
coef(iv_iq)["educ"]
```

**(a)** The 2SLS estimate is larger than the OLS estimate.

**(b)** One reason OLS might underestimate the return to education is the measurement error in estimating years of schooling, which would bias the coefficient toward zero. Additionally, the IV estimate captures the effect for compliers (people whose education is influenced by their distance to a college) who may have higher marginal returns to education.

**(c)** The 2SLS estimate is more credible as a causal estimate because it uses exogenous variation in education caused by proximity to a college. However, its validity still depends on the exclusion restriction holding true.

4.5

**(a)** Including IQ reduces the OLS coefficient on education from 0.0715 to 0.0676 which indicates that some of the OLS estimate reflects differences in ability.

**(b)** The 2SLS estimate also decreases when IQ is included, from 0.1249 to 0.1094, but it is substantially larger than the OLS estimate.

**(c)** This comparison suggests that ability bias explains some parts, but not all, of the difference between OLS and IV estimates. The remaining differences is likely due to measurement error in education or heterogeneous returns, rather than some other omitted variable by itself.
